Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2022-04-06 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex

\Abstract{
This is a tutorial paper on imputing incomplete multilevel data with
\pkg{mice}. Footnotes in the current version show work in progress/under
construction. The last section is not part of the manuscript, but purely
for reminders. We aim to submit at JSS, so there is no word count limit
(``There is no page limit, nor a limit on the number of figures or
tables'').
}

@@

In many contemporary data analysis efforts, some form of hierarchical or
clustered data structures are recorded. In the simplest case, such a
structure entails the nesting of units within clusters (e.g., students
within school classes). More complex clustered structures may occur when
there are multiple hierarchical levels (e.g., patients within hospitals
within regions or countries), or when the clustering is non-nested
(e.g., data from diverse settings and populations within large
databases). The clustered structure of multilevel data should be taken
into account when developing analysis models: 1) for the simple reason
that groups of observations share some common variance, and 2) because
ignoring multilevel structures can be harmful to the statistical
inferences and introduce bias in estimators \citep{hox17}. There are
many names for models that take clustering into account. Some popular
examples are `multilevel models', `hierarchical models', `mixed effect
models' and `random effect models'. Table \ref{tab:def} provides an
overview of some key concepts in multilevel modeling.

\begin{table}[tb]
\caption{Concepts in multilevel modeling}
\label{tab:def}
\centering
\begin{tabular}{ll}
\hline
\textbf{Concept} & \textbf{Details}   \\
\hline
ICC                 & The variability due to clustering is often measured by means of the \\
                    & intraclass coefficient (ICC). The ICC can be seen as the percentage \\
                    & of variance that can be attributed to the cluster-level, where a high \\
                    & ICC would indicate that a lot of variability is due to the cluster \\
                    & structure. \\
Random intercept    & Multilevel models typically accommodate for variability by including \\
                    & a separate group mean for each cluster. In addition to random \\
                    & intercepts, multilevel models can also include random effects and \\
                    & heterogeneous residual error variances across clusters [see e.g. \\
                    & @gelm06, @hox17 and @jong21]. \\
\hline
\end{tabular}
\end{table}

@@

The process of analyzing multilevel data is further complicated when not
all data entries are observed. Just as with single level data,
missingness may occur at the unit level. But with multiple levels of
data comes the potential for clustered missingness. Therefore,
incomplete multilevel data can be categorized into two general patterns:
systematic missingness and sporadic missingness \citep{resc13}.
Systematic missingness implies that one or more variables are never
observed in a certain cluster. With sporadic missingness there may be
observed data for some but not all units in a cluster
\citep{buur18, jola18}. We have visualized this difference in Figure 1,
which shows an \(n \times p\) set \(\mathbf{X} = X_1, \dots, X_p\), with
\(n\) units distributed over \(N\) clusters and \(p\) variables. Column
\(X_1\) is completely observed, column \(X_2\) is systematically missing
in cluster 2, and column \(X_3\) is sporadically missing. To analyze
these incomplete data, we have to take the nature of the missingness and
the cluster structure into account. For example, the sporadic
missingness in \(X_3\) could be easily amended if this would be a
cluster-level variable (and thus constant within clusters). We could
then just extrapolate the true (but missing) value of \(X_3\) for unit 1
from unit 2, and the value for unit 4 from unit 3. If \(X_3\) would
instead be a unit-level variable (which may vary within clusters), we
could not just recover the unobserved `truth', but would need to use
some kind of missing data method, or discard the incomplete units
altogether (i.e., complete case analysis). Complete case analysis can
however introduce bias in statistical inferences and lowers statistical
power. Further, with the systematic missingness in \(X_2\), it would be
impossible to fit a multilevel model without accommodating the
missingness in some way. Complete case analysis in that case would mean
excluding the entire cluster from the analyses. The wrong choice of
missing data handling method can thus be extremely harmful to the
inferences.

@@

Since excluding observations is not a desirable workflow, the
missingness in multilevel data should be accommodated \emph{before} or
\emph{within} the analysis of scientific interest. In this paper, we
focus on the former approach: imputing (i.e., filling in) the missing
data with plausible values, whereafter the completed data may be
analyzed as if it were completely observed. Imputation separates the
missing data problem from the scientific problem, which makes the
missing data strategy very generic and popular. If each missing value is
replaced multiple times, the resulting inferences may validly convey the
uncertainty due to missingness \citep[c.f.][]{rubi87}. The \proglang{R}
package \pkg{mice} has become the de-facto standard for imputation by
chained equations, which iteratively solves the missingness on a
variable-by-variable basis. \pkg{mice} is known to yield valid
inferences under many different missing data circumstances
\citep{buur18}. In this paper, we will discuss how to use \pkg{mice} in
the context of multilevel data.

@@

This papers serves as a tutorial for imputing incomplete multilevel data
with \pkg{mice}. We provide practical guidelines and code snippets for
different missing data situations, including missing not at random
(MNAR) mechanisms \citep[where the probability to be missing depends on
unrecorded information, making the missingness
non-ignorable,][]{rubi76, meng94}. For reasons of brevity, we focus on
imputation by chained equations wit \pkg{mice} exclusively\footnote{Note
  that the alternative, joint modeling imputation for multilevel data or
  \pkg{jomo} \citet{jomo}, has been implemented in \pkg{mice} as well
  but is outside the scope of this tutorial.}. Other useful resources
for the analysis of incomplete multilevel data include the \proglang{R}
packages \pkg{mitml}, \pkg{miceadds}, and \pkg{mdmb}, and empirical work
by \citet{audi18} and \citet{grun18}. Please note that this tutorial
paper assumes a basic level of knowledge on multilevel
models.\footnote{Note to self: We're providing an overview of
  implementations. It's up-to the reader to decide which multilevel
  strategy suits their data. We won't go into detail for the different
  methods (and equations). This paper is just a software tutorial, so
  we'll keep it practical.} Assumed knowledge also includes the use of
the `piping operator', \texttt{\%\textgreater{}\%}, adopted from the
\pkg{magrittr} package, and the \pkg{lme4} notation for multilevel
models.\footnote{TODO: Add environment info, seed and version number(s)
  somewhere!}

We illustrate how to impute incomplete multilevel data by means of three
case studies:

\begin{itemize}
\tightlist
\item
  \texttt{popmis} from the \pkg{mice} package \citep[simulated data on
  perceived popularity, \(n = 2,000\) pupils across \(N = 100\)
  schools,][]{mice};
\item
  \texttt{hiv} from the \pkg{GJRM} package \citep[simulated data on HIV
  diagnoses, \(n = 6,416\) patients across \(N = 9\) regions,][]{GJRM};
\item
  \texttt{impact} from the \pkg{metamisc} package \citep[empirical data
  on traumatic brain injuries, \(n = 11,022\) patients across \(N = 15\)
  studies,][]{metamisc}.
\end{itemize}

For each of these datasets, we will discuss the nature of the
missingness, choose one or more imputation models and evaluate the
imputed data, but we will also highlight one specific aspect of the
imputation workflow. With the \texttt{popmis} data, we show how (and how
not) to develop an imputation model. With the \texttt{hiv} data we focus
on extending the imputation model to include Heckman-type
selection-inclusion methods. With the \texttt{impact} data we provide an
example of multivariate missingness in real-world data. Together, this
should give enough scaffolding for applied researchers who are faced
with incomplete multilevel data.\footnote{TODO: Add notation paragraph
  or `translation table' linking multilevel equations to \pkg{lme4}
  formulas. Use betas instead of gamma's and mu's. Add interpretation of
  values in predictormatrix (-2 for the cluster variable, 2 for random
  effects). Add ICC and congeneality here as well. And make missingness
  mechanism table as well.}

Set-up the R environment and load the necessary packages:

\begin{CodeChunk}
\begin{CodeInput}
R> set.seed(2022)
R> library(mice)
R> library(ggmice)
R> library(ggplot2)
R> library(dplyr)
R> library(lme4)
R> library(mitml)
\end{CodeInput}
\end{CodeChunk}

\begin{center}
REVIEW UNTIL THIS POINT SVP
\end{center}

\hypertarget{case-study-i-how-not-to-impute}{%
\section{Case Study I: How (not) to
impute}\label{case-study-i-how-not-to-impute}}

In this section we'll go over the different steps involved with imputing
incomplete multilevel data. The data we're using is the \texttt{popmis}
dataset from the \texttt{mice} package. This is a simulated dataset with
pupils (\(n = 2000\)) clustered within schools (\(N = 100\)). In this
tutorial we'll use the following variables:

\begin{itemize}
\tightlist
\item
  \texttt{school}, school identification number (clustering variable);
\item
  \texttt{popular}, pupil popularity (self-rating between 0 and 10;
  unit-level);
\item
  \texttt{sex}, pupil sex (0=boy, 1=girl; unit-level);
\item
  \texttt{texp}, teacher experience (in years; cluster-level).
\end{itemize}

The analysis model corresponding to this dataset is multilevel
regression with random intercepts, random slopes and a cross-level
interaction. The outcome variable is \texttt{popular}, which is
predicted from the unit-level variable \texttt{sex} and the
cluster-level variable \texttt{texp}. The regression equation\footnote{add
  the `level notation' (Bryk and Raudenbush, 1992) and/or matrix
  notation (`linear mixed effects model'; Laird and Ware, 1982) too?}
and \texttt{lme4} notation for this model are

\[
\text{popular}_{ij} =
\gamma_{00} + 
\gamma_{10} \text{ sex}_{ij} + 
\gamma_{01} \text{ texp}_{j} + 
\gamma_{11} \text{ texp}_{j} \times \text{sex}_{ij} + 
u_{0j} + 
u_{1j} \text{ sex}_{ij} + 
e_{ij} \\
\]

\[
\texttt{popular} \sim  \texttt{1 + sex + texp + sex:texp + (1 + sex | school)}
\]

Since the data is simulated and the missingness is induced, we can
compare our inferences after imputation to the true complete data. The
data is created in such a way that the clustering variable
\texttt{school} explains quite some variance in the outcome variable
\texttt{popular}. We express this using the intraclass correlation, ICC
\(=\) 0.58. We'll evaluate the ICC after each missing data strategy, and
compare the estimated fixed effects:

\begin{CodeChunk}
\begin{CodeOutput}
                 Estimate with 95% CI
1 (Intercept)  3.314 [ 2.998,  3.629]
2         sex  1.330 [ 1.069,  1.590]
3        texp  0.110 [ 0.090,  0.130]
4    sex:texp -0.034 [-0.051, -0.017]
\end{CodeOutput}
\end{CodeChunk}

\hypertarget{incomplete-data}{%
\subsubsection{Incomplete data}\label{incomplete-data}}

Load the data into the environment and select the relevant variables:

\begin{CodeChunk}
\begin{CodeInput}
R> popmis <- popmis[, c("school", "popular", "sex", "texp")] 
\end{CodeInput}
\end{CodeChunk}

Plot the missing data pattern:

\begin{CodeChunk}
\begin{CodeInput}
R> plot_pattern(popmis)
\end{CodeInput}
\begin{figure}

{\centering \includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_pat-1}

\caption[Missing data pattern in the popularity data]{Missing data pattern in the popularity data}\label{fig:pop_pat}
\end{figure}
\end{CodeChunk}

The missingness is univariate and sporadic, which is illustrated in the
missing data pattern in Figure \ref{fig:pop_pat}. The ICC in the
incomplete data is 0.56. This tells us that the multilevel structure of
the data should probably be taken into account. If we don't, we'll may
end up with incorrect imputations, biasing the effect of the clusters
towards zero.

Plot the correlations in the incomplete data:

\begin{CodeChunk}
\begin{CodeInput}
R> plot_corr(popmis)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop-corr-1} \end{center}

@@

To develop the best imputation model for the incomplete variable
\texttt{popular}, we need to know whether the missingness depends on the
observed values of other variables. We'll highlight one other variable
to illustrate, but ideally one would inspect all relations. The
questions we'll ask are: `Does the missing data of pupil popularity
(\texttt{popular}) depend on observed teacher popularity
(\texttt{texp})?'. This can be evaluated statistically, but visual
inspection usually suffices. We'll make a histogram of \texttt{texp}
separately for the pupils with known popularity and missing popularity.

Plot the histogram for teacher experience conditional on the missingness
indicator of \texttt{popular}:

\begin{CodeChunk}
\begin{CodeInput}
R> ggmice(popmis, aes(texp)) +
+   geom_histogram(fill = "white") +
+   facet_grid(. ~ is.na(popular), labeller = label_both)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop-hist-1} \end{center}

TODO: think about what is a meaningful rule of thumb to signal that the
user should be worried?

This shows us that there are no apparent differences in the distribution
of \texttt{texp} depending on the missingness indicator of
\texttt{popular} (t = -0.873, p = 0.383).

\hypertarget{complete-case-analysis-not-recommended}{%
\subsubsection{Complete case analysis (not
recommended)}\label{complete-case-analysis-not-recommended}}

Complete case analysis ignores the observations with missingness
altogether, which lowers statistical power and may even introduce bias
in MCAR situations.

\hypertarget{imputation-ignoring-the-cluster-variable-not-recommended}{%
\subsubsection{Imputation ignoring the cluster variable (not
recommended)}\label{imputation-ignoring-the-cluster-variable-not-recommended}}

The first imputation model that we'll use is likely to be invalid. We do
\emph{not} use the cluster identifier \texttt{school} as imputation
model predictor. With this model, we ignore the multilevel structure of
the data, despite the high ICC. This assumes exchangeability between
units. We include it purely to illustrate the effects of ignoring the
clustering in our imputation effort. We'll use the default imputation
methods in \texttt{mice()} (predictive mean matching to impute the
continuous variables and logistic regression to impute binary
variables).

Create a methods vector and predictor matrix for \texttt{popular}, and
make sure \texttt{school} is not included as predictor:

\begin{CodeChunk}
\begin{CodeInput}
R> meth <- make.method(popmis) # methods vector
R> pred <- quickpred(popmis)   # predictor matrix
R> plot_pred(pred)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop-ignored-pred-1} \end{center}

Impute the data, ignoring the cluster structure:

\begin{CodeChunk}
\begin{CodeInput}
R> imp_ignored <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
\end{CodeInput}
\end{CodeChunk}

TODO: remove the broom.mixed output, use mitml only

Analyze the imputations:

\begin{CodeChunk}
\begin{CodeInput}
R> fit_ignored <- imp_ignored %>% 
+   with(lme4::lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school))) 
R> testEstimates(as.mitml.result(fit_ignored), var.comp = TRUE)
\end{CodeInput}
\begin{CodeOutput}

Call:

testEstimates(model = as.mitml.result(fit_ignored), var.comp = TRUE)

Final parameter estimates and inferences obtained from 5 imputed data sets.

             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
(Intercept)     3.354     0.245    13.704     6.831     0.000     3.260     0.813 
sex             1.247     0.332     3.759     6.154     0.009     4.160     0.849 
texp            0.110     0.017     6.320     6.084     0.001     4.286     0.852 
sex:texp       -0.027     0.022    -1.225     5.914     0.267     4.631     0.862

                            Estimate 
Intercept~~Intercept|school    0.177 
sex~~sex|school                0.232 
Intercept~~sex|school         -0.043 
Residual~~Residual             0.670 
ICC|school                     0.209

Unadjusted hypothesis test as appropriate in larger samples.
\end{CodeOutput}
\end{CodeChunk}

\begin{CodeChunk}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_ignored_eval-1} \end{center}

\hypertarget{imputation-with-the-cluster-variable-as-predictor-not-recommended}{%
\subsubsection{Imputation with the cluster variable as predictor (not
recommended)}\label{imputation-with-the-cluster-variable-as-predictor-not-recommended}}

We'll now use \texttt{school} as a predictor to impute all other
variables. This is still not recommended practice, since it only works
under certain circumstances and results may be biased
\citep{drec15, ende16}. But at least, it includes some multilevel
aspect. This method is also called `fixed cluster imputation', and uses
N-1 indicator variables representing allocation of N clusters as a fixed
factor in the model \citep{reit06, ende16}. Colloquially, this is
`multilevel imputation for dummies'.

Add: doesn't work with syst missing (only sporadically). There's some
pro's and con's. May not differ much if the number of clusters is low.

The more the random effects are of interest, the more you need ml
models.

\begin{CodeChunk}
\begin{CodeInput}
R> # adjust the predictor matrix
R> pred["popular", "school"] <- 1 
R> plot_pred(pred)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_predictor-1} \end{center}

\begin{CodeInput}
R> # impute the data, cluster as predictor
R> imp_predictor <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_predictor_eval-1} \end{center}

Now, we can clearly see that the imputed values of \texttt{texp} are
higher than the observed values, which is in line with right-tailed MAR.

The ICCs are way more in line with the ICCs in the incomplete data. But
this is a quick and dirty way of imputing multilevel data. We
\emph{should} be using a multilevel model.

\hypertarget{imputation-with-random-effects}{%
\subsubsection{Imputation with random
effects}\label{imputation-with-random-effects}}

With \texttt{2l.norm} we impute the outcome with a multilevel model
assuming random slopes for each variable in the imputation model and
homogeneous within-cluster variance.

``Van Buuren (2011) considered the homoscedastic linear mixed model as
invalid for imputing incomplete predictors, and investigated only the
2l.norm method, which allows for heterogeneous error variances''
\citep{buur18}.

\begin{CodeChunk}
\begin{CodeInput}
R> # adjust the predictor matrix
R> pred["popular", ] <- c(school = -2, popular = 0, sex = 2, texp = 2) 
R> plot_pred(pred) 
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_norm-1} \end{center}

\begin{CodeInput}
R> meth <- make.method(popmis)
R> meth["popular"] <- "2l.pmm"
R> imp_pmm_2l <-
+   mice(
+     popmis %>% mutate(school = as.integer(school)),
+     pred = pred,
+     meth = meth,
+     maxit = 1,
+     print = FALSE
+   )
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_norm_eval-1} \end{center}

\hypertarget{imputation-with-random-effects-and-heterogeneity}{%
\subsubsection{Imputation with random effects and
heterogeneity}\label{imputation-with-random-effects-and-heterogeneity}}

This method assumes random slopes for each variable in the imputation
model. In contrast to \texttt{2l.norm} this method allows a
cluster-specific residual error variance.

\begin{CodeChunk}
\begin{CodeInput}
R> pred["popular", ] <- c(-2, 2, 1, 2)
R> meth <- c("", "2l.pan", "", "")
R> imp_pan_2l <-
+   mice(
+     popmis %>% mutate(school = as.integer(school)),
+     pred = pred,
+     meth = meth,
+     maxit = 1,
+     print = FALSE
+   )
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_pan_eval-1} \end{center}

\hypertarget{how-to-handle-non-random-selection-case-study-ii-hiv}{%
\section{How to handle non-random selection (Case study II:
HIV)}\label{how-to-handle-non-random-selection-case-study-ii-hiv}}

Data are simulated and included in the \texttt{GJRM} package. We will
use the following variables:

\begin{itemize}
\tightlist
\item
  \texttt{region} Cluster variable,
\item
  \texttt{hiv} HIV diagnosis (0=no, 1=yes),
\item
  \texttt{age} Age of the patient,
\item
  \texttt{marital} Marital status,
\item
  \texttt{condom} Condom use during last intercourse,
\item
  \texttt{smoke} Smoker (levels; inclusion restriction variable).
\end{itemize}

The imputation of these date is based on the toy example from
\href{https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R}{IPDMA
Heckman Github repo}.

\begin{CodeChunk}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/hiv-1} \end{center}

From the missing data pattern we see that we can set \texttt{maxit} to
1, since there is only one variable with missingness.

The inclusion restriction variable should be a predictor of the the
actual value of the variable of interest, but \emph{not} of missingness
indicator for the variable of interest. In this case, the data were
simulated to adhere to this requirement. Namely, \(\beta_{smoke}\) =
-0.064, 95\% CI {[}-0.256, 0.126{]} for the analysis model
(\texttt{formula\ =\ hiv\ \textasciitilde{}\ .}), and \(\beta_{smoke}\)
= -0.265, 95\% CI {[}-0.422, -0.11{]} for the selection model
(\texttt{formula\ =\ is.na(hiv)\ \textasciitilde{}\ .}). This means the
assumptions for the Heckman-type selection model are met.

\hypertarget{how-to-handle-multivariate-missingness-case-study-iii-impact}{%
\section{How to handle multivariate missingness (Case study III:
IMPACT)}\label{how-to-handle-multivariate-missingness-case-study-iii-impact}}

\texttt{impact} is traumatic brain injury data with patients,
\(n = 11022\), clustered in studies, \(N = 15\). With the following 11
variables:

The analysis model for this dataset is a prediction model with
\texttt{mort} as the outcome.\footnote{Look at analysis model, maybe
  copy from GREAT data example e.g., adjusted prognostic effect of
  \texttt{ct} on unfortunate outcomes, we just want to know the adjusted
  odds ratio for \texttt{ct}. Add something about systematically missing
  data here.} The data is already imputed (Steyerberg et al, 2008), so
we've induced missingness again based on the missingness in the original
data.

\[
\text{mort}_{ij} =
\gamma_{00} + 
\gamma_{01} \text{ type}_{j} + 
\gamma_{10} \text{ age}_{ij} + 
\gamma_{10} \text{ moter\_score}_{ij} + 
\gamma_{10} \text{ pupil}_{ij} + 
\gamma_{10} \text{ ct}_{ij} + \\
u_{0j} + 
u_{1j} \text{ age}_{ij} + 
u_{1j} \text{ moter\_score}_{ij} +  
u_{1j} \text{ pupil}_{ij} +  
u_{1j} \text{ ct}_{ij} +  
e_{ij} \\
\]

>>>>

\Abstract{
This is a tutorial paper on imputing incomplete multilevel data with
\pkg{mice}. Footnotes in the current version show work in progress/under
construction. The last section is not part of the manuscript, but purely
for reminders. We aim to submit at JSS, so there is no word count limit
(``There is no page limit, nor a limit on the number of figures or
tables''). {[}Just adding some text to get a better guess of what the
actura abstract will look like: Lorem ipsum dolor sit amet, consectetur
adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore
magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco
laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor
in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla
pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa
qui officia deserunt mollit anim id est laborum.{]}
}

@@

In many contemporary data analysis efforts, some form of hierarchical or
clustered data structures are recorded. In the simplest case, such a
structure entails the nesting of units within clusters (e.g., students
within school classes). More complex clustered structures may occur when
there are multiple hierarchical levels (e.g., patients within hospitals
within regions or countries), or when the clustering is non-nested
(e.g., electronic health record data from diverse settings and
populations within large databases). The clustered structure of
multilevel data should be taken into account when developing analysis
models: 1) for the simple reason that groups of observations share some
common variance, and 2) because ignoring multilevel structures can be
harmful to the statistical inferences and introduce bias in estimators
\citep{hox17}. There are many names for models that take clustering into
account. Some popular examples are `multilevel models', `hierarchical
models', `mixed effect models' and `random effect models'. Table
\ref{tab:clus} provides an overview of some key concepts in multilevel
modeling.

\begin{table}[tb]
\caption{Concepts in multilevel methods}
\label{tab:clus}
\centering
\begin{tabular}{ll}
\hline
\textbf{Concept} & \textbf{Details}   \\
\hline
ICC                 & The variability due to clustering is often measured by means of the \\
                    & intraclass coefficient (ICC). The ICC can be seen as the percentage \\
                    & of variance that can be attributed to the cluster-level, where a high \\
                    & ICC would indicate that a lot of variability is due to the cluster \\
                    & structure. \\
Random intercept    & Multilevel models typically accommodate for variability by including \\
                    & a separate group mean for each cluster. In addition to random \\
                    & intercepts, multilevel models can also include random effects and \\
                    & heterogeneous residual error variances across clusters [see e.g. \\
                    & @gelm06, @hox17 and @jong21]. \\
\hline
\end{tabular}
\end{table}

@@

The process of analyzing multilevel data is further complicated when not
all data entries are observed. Just as with single level data,
missingness may occur at the unit level. But with multiple levels of
data comes the potential for clustered missingness. Therefore,
incomplete multilevel data can be categorized into two general patterns:
systematic missingness and sporadic missingness \citep{resc13}.
Systematic missingness implies that one or more variables are never
observed in a certain cluster. With sporadic missingness there may be
observed data for some but not all units in a cluster
\citep{buur18, jola18}. We have visualized this difference in Figure 1,
which shows an \(n \times p\) set \(\mathbf{X} = X_1, \dots, X_p\), with
\(n\) units distributed over \(N\) clusters and \(p\) variables.

@@

Column \(X_1\) in Figure 1 is completely observed, column \(X_2\) is
systematically missing in cluster 2, and column \(X_3\) is sporadically
missing. To analyze these incomplete data, we have to take the nature of
the missingness and the cluster structure into account. For example, the
sporadic missingness in \(X_3\) could be easily amended if this would be
a cluster-level variable (and thus constant within clusters). We could
then just extrapolate the true (but missing) value of \(X_3\) for unit 1
from unit 2, and the value for unit 4 from unit 3. If \(X_3\) would
instead be a unit-level variable (which may vary within clusters), we
could not just recover the unobserved `truth', but would need to use
some kind of missing data method, or discard the incomplete units
altogether (i.e., complete case analysis). Complete case analysis can
however introduce bias in statistical inferences and lowers statistical
power. Further, with the systematic missingness in \(X_2\), it would be
impossible to fit a multilevel model without accommodating the
missingness in some way. Complete case analysis in that case would mean
excluding the entire cluster from the analyses. The wrong choice of
missing data handling method can thus be extremely harmful to the
inferences.

A key characteristic of the missing data to take into account in
analyses is the mechanism behind the missingness. We distinguish between
MCAR, MAR and MNAR in theory (see Table \ref{tab:miss}), but in practice
this distinction is less clear. Since the essence of the true
non-response mechanism may not be known, it is generally inferred or
assumed to be ignorable (i.e., MCAR or MAR). {[}TODO: add that this
assumption may not always be valid, especially with modern types of big
dasta sources.{]} Depending on the actual missingness-generating
mechanism, missing data handling strategies may be more or less
suitable, see e.g., \citet{yuce08} and \citet{hox15}.

\begin{table}[tb]
\caption{Concepts in missing data methods}
\label{tab:miss}
\centering
\begin{tabular}{ll}
\hline
\textbf{Concept} & \textbf{Details}   \\
\hline
MCAR    & Missing Completely At Random, where the probability to be missing is equal \\
& across all data entries \\
MAR     & Missing At Random, where the probability to be missing depends on observed \\
& information \\
MNAR    & Missing Not At Random (MNAR), where the probability to be missing \\
& depends on unrecorded information, making the missingness non-ignorable \\
& [@rubi76; @meng94]. \\
& [TODO: add congeniality.] \\
\hline
\end{tabular}
\end{table}

Since excluding observations is not a desirable workflow, the
missingness in multilevel data should be accommodated \emph{before} or
\emph{within} the analysis of scientific interest. In this paper, we
focus on the former approach: imputing (i.e., filling in) the missing
data with plausible values, whereafter the completed data may be
analyzed as if it were completely observed. Imputation separates the
missing data problem from the scientific problem, which makes the
missing data strategy very generic and popular. If each missing value is
replaced multiple times, the resulting inferences may validly convey the
uncertainty due to missingness \citep[c.f.][]{rubi76}. The \proglang{R}
package \pkg{mice} has become the de-facto standard for imputation by
chained equations, which iteratively solves the missingness on a
variable-by-variable basis. \pkg{mice} is known to yield valid
inferences under many different missing data circumstances
\citep{buur18}. In this paper, we will discuss how to use \pkg{mice} in
the context of multilevel data.

{[}TODO: clarify why clustering is relevant during imputation, and why
this exposes the need for specialized imputation methods and more
attention during their implementation (``thou shall not simply run
\texttt{mice()} on any incomplete dataset''). And add overview of
possible predictor matrix values.{]}

@@

This papers serves as a tutorial for imputing incomplete multilevel data
with \pkg{mice} in \proglang{R}. We provide practical guidelines and
code snippets for different missing data situations, including
non-ignorable mechanisms. For reasons of brevity, we focus on multilevel
imputation by chained equations with \pkg{mice} exclusively; other
imputation methods and packages (e.g., \pkg{jomo} and \pkg{mdmb}) are
outside the scope of this tutorial. Assumed knowledge includes basic
familiarity with multilevel imputation \citep[see e.g.][ and
\citet{grun18}]{audi18} and the \pkg{lme4} notation for multilevel
models (see Table \ref{tab:mod}).

\begin{table}[tb]
\caption{Concepts in missing data methods}
\label{tab:mod}
\centering
\begin{tabular}{ll}
\hline
\textbf{Concept} & \textbf{Details}   \\
\hline
& [TODO: explain \pkg{lme4} notation here] \\
\hline
\end{tabular}
\end{table}

\hypertarget{case-study}{%
\section{Case study}\label{case-study}}

We illustrate how to impute incomplete multilevel data by means of a
case study: \texttt{impact} from the \pkg{metamisc} package
\citep[empirical data on traumatic brain injuries, \(n = 11,022\)
patients across \(N = 15\) studies,][]{metamisc}. \texttt{impact} is
traumatic brain injury data with patients, \(n = 11022\), clustered in
studies, \(N = 15\). With the following 11 variables:

{[}TODO: add more info about the complete data.{]} The analysis model
for this dataset is a prediction model with \texttt{mort} as the
outcome. In this tutorial we'll estimate the adjusted prognostic effect
of \texttt{ct} on unfortunate outcomes. The estimand is the adjusted
odds ratio for \texttt{ct}, after including \texttt{type}, \texttt{age}
\texttt{motor\_score} and \texttt{pupil} into the analysis model:

\begin{CodeChunk}
\begin{CodeInput}
R> mod <- mort ~ 1 + type + age + motor_score + pupil + ct + (1 | name)
\end{CodeInput}
\end{CodeChunk}

Note that variables \texttt{hypots}, \texttt{hypox}, \texttt{tsah} and
\texttt{edh} are not part of the analysis model, and may thus serve as
auxiliary variables for imputation.

\hypertarget{setup}{%
\subsection{Setup}\label{setup}}

{[}TODO: Add environment info, seed and version number(s) somewhere.{]}
Set up the R environment and load the necessary packages:

\begin{CodeChunk}
\begin{CodeInput}
R> set.seed(2022)
R> library(mice)         # for imputation
R> library(ggmice)       # for visualization
R> library(ggplot2)      # for visualization
R> library(dplyr)        # for data wrangling
R> library(lme4)         # for multilevel modeling
R> library(broom.mixed)  # for multilevel estimates
R> library(mitml)        # for multilevel pooling
\end{CodeInput}
\end{CodeChunk}

The \texttt{impact} data included in the \pkg{metamisc} package is a
complete data set. The original data has already been imputed once
(Steyerberg et al, 2008). For the purpose of this tutorial we have
induced missingness (mimicking the missing data in the original data set
before imputation). The resulting incomplete data can be accessed from
\href{https://zenodo.com}{zenodo link to be created}.

Load the complete and incomplete data into the R workspace:

\begin{CodeChunk}
\begin{CodeInput}
R> data("impact", package = "metamisc")
R> dat <- read.table("link/to/the/data.txt")
\end{CodeInput}
\end{CodeChunk}

We will use the following estimates as comparative truth in this
tutorial {[}TODO: make this a table or forest plot instead, to see if
there is heterogeneity in the association of \texttt{ct} with
\texttt{mort}{]}:

\begin{CodeChunk}
\begin{CodeInput}
R> fit <- glmer(mod, family = "binomial", data = impact) 
R> tidy(fit, conf.int = TRUE, exponentiate = TRUE) 
\end{CodeInput}
\begin{CodeOutput}
# A tibble: 11 x 9
   effect group term  estimate std.error statistic    p.value conf.low conf.high
   <chr>  <chr> <chr>    <dbl>     <dbl>     <dbl>      <dbl>    <dbl>     <dbl>
 1 fixed  <NA>  (Int~    0.101   0.0173     -13.4   4.94e- 41   0.0723     0.141
 2 fixed  <NA>  type~    0.713   0.123       -1.96  5.01e-  2   0.509      1.00 
 3 fixed  <NA>  age      1.03    0.00165     20.2   2.13e- 90   1.03       1.04 
 4 fixed  <NA>  moto~    0.553   0.0386      -8.50  1.95e- 17   0.482      0.634
 5 fixed  <NA>  moto~    0.405   0.0289     -12.7   8.43e- 37   0.352      0.466
 6 fixed  <NA>  moto~    0.275   0.0202     -17.6   1.67e- 69   0.239      0.318
 7 fixed  <NA>  pupi~    3.73    0.230       21.4   2.09e-101   3.31       4.21 
 8 fixed  <NA>  pupi~    1.87    0.133        8.80  1.36e- 18   1.63       2.15 
 9 fixed  <NA>  ctIII    2.25    0.157       11.6   5.12e- 31   1.96       2.58 
10 fixed  <NA>  ctIV~    2.30    0.136       14.0   9.47e- 45   2.05       2.58 
11 ran_p~ name  sd__~    0.277  NA           NA    NA          NA         NA    
\end{CodeOutput}
\end{CodeChunk}

{[}TODO: add ICC before/after imputation and interpret: This tells us
that the multilevel structure of the data should probably be taken into
account. If we don't, we'll may end up with incorrect imputations,
biasing the effect of the clusters towards zero.{]}

{[}TODO: add descriptive statistics of the complete and incomplete
data.{]}

\hypertarget{missingness}{%
\subsection{Missingness}\label{missingness}}

To explore the missingness, it is wise to look at the missing data
pattern:

\begin{CodeChunk}
\begin{CodeInput}
R> plot_pattern(dat, rotate = TRUE)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pattern-1} \end{center}

@@

This shows\ldots{} {[}TODO: fill in that we need to impute \texttt{ct}
and \texttt{pupil}.{]}

To develop the best imputation model, we need to investigate the
relations between the observed values of the incomplete variables and
the observed values of other variables, and the relation between the
missingness indicators of the incomplete variables and the observed
values of the other variables. To see whether the missingness depends on
the observed values of other variables, we\ldots{} {[}TODO: fill in that
we can test this statistically or use visual inspection (e.g.~a
histogram faceted by the missingness indicator).{]}
